Type Classes for Mathematics in Type Theory

### 2010-06-30 jar.tex
Many lay-out, text changes

\begin{document}
\title{Type Classes for Mathematics in Type Theory}
\author{Bas Spitters \and Eelis van der Weegen} % For MSCS \\ Radboud University Nijmegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
The introduction of first-class type classes in the Coq system calls for re-examination of the basic interfaces used for mathematical formalization in type theory.
% especially since existing solutions fall short from a proof engineering perspective.
We present a new set of foundational type classes for mathematics and take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible. Thus, we address both traditional proof engineering challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.

@@

For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation (e.g.\ rewriting).

Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a. telescopes), packed classes~\cite{Packed}, and occasionally modules. We present a new approach based entirely on Coq's new type class mechanism, and show how it can be used to make fully `unbundled' predicate-representations of algebraic structures practical to work with.

Since we intend to use this development as a basis for constructive analysis with practical certified exact real arithmetic, an additional objective and motivation in our design is to facilitate \emph{efficient} computation. In particular, we want to be able to effortlessly swap implementations of number representations. Doing this requires that we have clean abstract interfaces, and mathematics tells us what these should look like: we represent $\N$, $\Z$, and $\Q$ as \emph{interfaces} specifying an initial semiring, an initial ring, and a field of integral fractions, respectively.

To express these elegantly and without duplication, our development\footnote{The sources are available
at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of parts of category theory and multi-sorted universal algebra, all expressed using type classes for optimum effect.

@@

\paragraph{Outline}
In section~\ref{preliminaries} we briefly describe the Coq system and some of its facilities that are of particular interest to us (most notably its implementation of type classes). Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures. In section~\ref{predicateclasses} we show how type classes can make the use of `unbundled' purely predicate based interfaces for abstract structures practical. Next, in section~\ref{classes}, we discuss our algebraic hierarchy implemented using such predicate classes. In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers. In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}. In section~\ref{sequences} we hint at an interface for sequences but describe how a limitation in the Coq system makes its use problematic. We end with conclusions in section~\ref{conclusions}.

@@

The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types; see~\cite{Coq,BC04}. In true Curry-Howard fashion, it can be thought of either as an excessively pure (if somewhat pedantic) functional programming language with an extremely expressive type system, or as a language for mathematical statements and proofs. We highlight some aspect of Coq that are of particular relevance to our development.

\paragraph{Types and propositions}

Propositions in Coq are types~\cite{ITT,CMCP}, which themselves have types called \emph{sorts}. Coq features a distinguished sort called \lstinline|Prop| that one may voluntarily choose to use as the sort for types representing propositions. The distinguishing feature of the \lstinline|Prop| sort is that Coq will not permit terms of non-\lstinline|Prop| types to depend on the values of inhabitants of \lstinline|Prop| types (that is, proof terms).

% As originally conceived~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
% them. Often we do \emph{not} want functions to depend on proofs. In Coq, this is the paradigmatic way of working. We are forbidden to extract information from proofs of \lstinline|Prop|s. Especially with the aid of the Program machinery, this allows us to flexibly work with simply typed program and their Hoare style correctness proofs.

Adopting this regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations. On a very practical level, this lets Coq safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.

A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.
% todo: process?:
%One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
%construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
%the [ ]-types are isomorphic to the type of sort \lstinline|Prop|. However,
%there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
%theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
%irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
%of Propositions as types seems necessary.
% Making Harrop formulas proof irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. An orthogonal and further reaching proposal is to replace the Coq type theory with the Calculus of implicit constructions with implicit Σ-types~\cite{miquel2001implicit,barras2008implicit,Bernardo}.

@@

The `native' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor \lstinline|eq_refl| of type \lstinline|Π (T: Type) (x: T), x ≡ x|, where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve \lstinline|=| for setoid equality, as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions.

We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object (for instance when identifying formal fractions of integers representing the same ratio). Rewriting a subterm using such an equality is permitted only if the subterm is an argument of a function that has been proven to \emph{respect} the equality. Such a function is called \emph{proper}, and propriety must be proved for each function in whose arguments we wish to enable rewriting.

Because the Coq type theory lacks quotient types (as it would make type checking undecidable), the conventional way of working with abstract structures in Coq is to always start with \emph{setoids} (aka Bishop sets): types equipped with an equivalence relation~\cite{Bishop67,Hofmann,Capretta}. As we will see in section ~\ref{univ}, this way of working by nature lends itself well to particularly simple definitions of notions such as quotient algebras.

@@

For an interesting (if fairly restrictive) alternative approach based on the idea of requiring unicity of representation for all objects so that Leibniz equality is always applicable, see the algebraic hierarchy part of the ssreflect~\cite{Packed} libraries.

\paragraph{Type classes}

Type classes have been a great success story in the Haskell functional programming language, as a means for organizing interfaces of abstract structures. The type class implementation that is now part of Coq provides a superset of their functionality, but implemented in a different way.

In Haskell and Isabelle, type classes and their instances are second class, in that they are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus. By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (`dictionaries'), instances are ordinary constants of these record types (registered as `hints' with the proof search machinery), class constraints are ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.

Thus, type classes in Coq are realized by relatively minor syntactic aids that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.

@@

Higher structures, too, can all be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in principle, in practice \emph{naive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.

@@

Manifest fields~\cite{DBLP:conf/types/Luo08} have been proposed and implemented in the Matita system to address exactly this problem. We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution (described in the next section) that avoids the problem altogether by using predicate-like type classes in place of bundled records.

If we revert back to the predicate formulation of relations, we \emph{could} still define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
\begin{lstlisting}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: relation er_carrier
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}
However, the same argument as before leads to the conclusion that \mbox{\lstinline|EquivalenceRelation|,} too, should be a predicate. Indeed, it would be rather strange for the interface of equivalence relations to differ qualitatively from the interface of reflexive relations.

Another attempt to recover some grouping might be to bundle the carrier with the relation into a (lawless) record, but this too hinders sharing: as soon as we try to define an algebraic structure with two reflexive relations on the same carrier, we need awkward hacks to establish equality between the carrier projections of two different (carrier, relation) bundles.

Even bundling just the operations of an algebraic structure together in a record (with the carrier as a parameter) does not work, as it again leads to the same problem when, for example, one attempts to define a hypothetical algebraic structure with two binary relations and a constant such that both binary relations form a monoid with the constant.

A second problem with bundling is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (coercions and notations can make this less painful). Further, if one tries to implement some semblance of sharing in a bundled representation, these projection paths additionally become non-canonical, and still more extensions have been proposed to address this symptom (e.g.\ coercion pullbacks~\cite{Hints}).

@@

Next, we turn to problems concerning theorem \emph{declaration}. As a reference point, our ideal will be the common mathematical vernacular, where one simply says:
\begin{quote}
Theorem L: For $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$.
\end{quote}
(Please forgive the admittedly silly statement; we are presently concerned only with the syntax.)

@@

\begin{lstlisting}
  Theorem L G e op {P: SemiGroup G e op}: Π x y z,
    equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op z y) x).
\end{lstlisting}
where e and the op's are filled in by instance resolution.

@@

Since \lstinline|CommutativeMonoid| indirectly includes a \lstinline|SemiGroup| field, which in turn includes a \lstinline|Equivalence| field, having a \lstinline|SemiRing| proof means having two distinct proofs that the equality relation is an equivalence. This kind of redundant knowledge (which arises naturally in mathematics) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs. We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the `right' proofs of properties. There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer to) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than `mere' convertibility) to recognize identical subterms.

Because predicate classes only provide contextual information and are insulated from the actual algebraic expressions, their instances can always be kept entirely opaque---only their existence matters. Together, these properties largely defuse an argument occasionally voiced against type classes concerning perceived unpredictability of instance resolution. While it is certainly true that in contexts with redundant information it can become hard to predict which instance of a predicate class will be found by proof search, this is simply a non-issue because it \emph{does not matter} which one is found. And for operational type classes the issue rarely arises because their instances are not nearly as abundant.

@@

The reader will note that we use names for properties like distributivity and absorption. Indeed, these are type classes as well (which is why we declare them with \lstinline|:>|). It has been our experience that almost any generic predicate worth naming is worth representing as a predicate type class, so that its proofs will be resolved as instances behind the scenes whenever possible. Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular. Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a `database' of properties previously proved. Much more so than the existing canonical structures facility, because canonical structures can only be used with bundled representations, which we argued against in section~\ref{bundling}.

The operational type classes used in \lstinline|SemiRing| for zero, one, multiplication and addition, are the same ones used by \lstinline|Ring| and \lstinline|Field| (not shown). Thus, the realization that a particular semiring is in fact a ring or field (etc.) has no bearing on how one refers to the operations in question, which is as it should be. The realization that a particular semigroup is part of a semiring \emph{does} call for a new (canonical) name, simply because of the need for disambiguation. The introduction of these additional names for the same operation is quite harmless in practice, because canonical names established by operational type class fields are just trivial identity functions anyway, so that in most contexts the distinction reduces away instantly.

@@

Some clarification is in order to explain the role of the \lstinline|Context| declaration of the two monoids. While it may look we are parameterizing \lstinline|Monoid_Morphism| on monoid-ness proofs (which would be a gross violation of our idiom), in fact it is only parameterized on the monoid \emph{components} declared through implicit generalization of the \lstinline|Monoid| declarations, because it only refers to those. Here, we use declarations of predicate class parameters merely as convenient shorthands to declare their components.

@@

Unfortunately, there is actually one annoying wrinkle here, which will also explain why, as the observant reader will have noticed, we do not register these two fields as instance resolution hints (by declaring them with \lstinline|:>|). What we really want these fields to express is ``\emph{if} in a certain context we know something to be a \lstinline|Monoid_Morphism|, \emph{then} realize that the source and target are \lstinline|Monoid|s''. However, the current instance resolution implementation has little support for this style of \emph{forward} reasoning, and is really primarily oriented on \emph{backward} reasoning: had we registered \lstinline|monmor_to| and \lstinline|monmor_from| as instance resolution hints, we would in fact be saying ``\emph{if} trying to establish that something is a \lstinline|Monoid|, \emph{then} try finding a \lstinline|Monoid_Morphism| to or from it'', which quickly degenerates into a wild goose chase. We will return to this point in section~\ref{conclusions}.

@@

In Coq, for a term \lstinline|F| to function as though it had two different types simultaneously (namely the object map and the arrow map), there must either (1) be coercions from the type of F to either function, or (2) F must be (coercible to) a single function that is able to consume both object and arrow arguments. In addition to not being supported by Coq, option (1) would violate our policy of leaving components unbundled. For (2), if it could be made to work at all (which is not clear at all), F would need a pretty egregious type considering that arrow types are indexed by objects, and that the type of the arrow map (namely \lstinline|Π x y, (x ⟶ y) → (F x ⟶ F y)|) must refer to the object map.

@@

To specify the natural numbers and the integers as initial objects in the categories of semirings and rings, respectively, definitions of these categories are needed. While one could define both of them manually, greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, for which \emph{varieties} can be defined generically (varieties are categories consisting of models for a fixed theory with homomorphisms between them).

To this end, we have formalized some of the theory of multisorted universal algebra and equational theories. We chose not to revive existing old (and somewhat stale) formalizations~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because an important aim for us has been to find out what level of elegance, convenience, and integration can be achieved by leveraging the state of the art in Coq facilities (of which type classes are the most important example).

@@

Taking these conversion functions for granted, we will also need a ``down-to-earth'' representation of the initiality arrows if we are to give a \lstinline|SemiRing|-based definition of the interface for natural numbers. Once again, we introduce an operational type class to represent this particular component:
\begin{lstlisting}
  Class NaturalsToSemiRing (A: Type) :=
    naturals_to_semiring: Π B `{RingMult B} `{RingPlus B} `{RingOne B}
      `{RingZero B}, A → B.
\end{lstlisting}
In order to use this with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \lstinline|SemiRing_Morphism|s, and builds an \lstinline|InitialArrow| instance out of it. This conversion instance in turn invokes another conversion function that translates concrete \lstinline|SemiRing_Morphism| proofs into univeral algebra \lstinline|Homomorphism|s instantiated with the semiring signature, which make up the arrows in the category.

@@

In our development we prove that the standard library's default rationals do indeed implement \lstinline|Rationals|, as do implementations of the \lstinline|QType| module interface. While the latter is rather ad-hoc from a theoretical perspective, it is nevertheless of great practical interest because it is used for the very efficient \lstinline|BigQ| rationals based on machine integers~\cite{machineintegers}. Hence, theory and programs developed on our \lstinline|Rationals| interface applies and can make immediate use of these efficient rationals. We plan to rebase the computable real number implementation in CoRN on this interface, precisely so that it may be instantiated with efficient implementations like these.

>>>>

\begin{document}
\title{Type Classes for Mathematics in Type Theory}
\author{Bas Spitters \and Eelis van der Weegen} % For MSCS \\ Radboud University Nijmegen}
\institute{Radboud University Nijmegen}
\date{today}
\maketitle
\begin{abstract}
The introduction of first-class type classes in the Coq system calls for re-examination of the basic interfaces used for mathematical formalization in type theory.
% especially since existing solutions fall short from a proof engineering perspective.
We present a new set of type classes for mathematics and take full advantage of their unique features to make practical a particularly flexible approach formerly thought infeasible. Thus, we address both traditional proof engineering challenges as well as new ones resulting from our ambition to build upon this development a library of constructive analysis in which abstraction penalties inhibiting efficient computation are reduced to a minimum.

@@

For the algebraic hierarchy---a critical component in any library of formalized mathematics---these challenges include structure inference, handling of multiple inheritance, idiomatic use of notations, and convenient algebraic manipulation (e.g.\ rewriting). Several solutions have been proposed for the Coq theorem prover: dependent records~\cite{DBLP:journals/jsc/GeuversPWZ02} (a.k.a. telescopes), packed classes~\cite{Packed}, and occasionally modules. We present a new approach based entirely on Coq's new type class mechanism, and show how it can be used to make fully `unbundled' predicate-representations of algebraic structures practical to work with.

Since we intend to use this development as a basis for constructive analysis with practical certified exact real arithmetic, an additional objective and motivation in our design is to facilitate \emph{efficient} computation. In particular, we want to be able to effortlessly swap implementations of number representations. Doing this requires that we have clean abstract interfaces, and mathematics tells us what these should look like: we represent $\N$, $\Z$, and $\Q$ as \emph{interfaces} specifying an initial semiring, an initial ring, and a field of integral fractions, respectively. To express these interfaces elegantly and without duplication, our development\footnote{The sources are available
at:~\url{http://www.eelis.net/research/math-classes/}} includes an integrated formalization of parts of category theory and multi-sorted universal algebra, all expressed using type classes for optimum effect.

@@

\paragraph{Outline.}
In section~\ref{preliminaries} we briefly describe the Coq system and its implementation of type classes. Then, in section~\ref{bundling}, we give a very concrete introduction to the issue of \emph{bundling}, arguably the biggest design dimension when building interfaces for abstract structures. In section~\ref{predicateclasses} we show how type classes can make the use of `unbundled' purely predicate based interfaces for abstract structures practical. Next, in section~\ref{classes}, we discuss our algebraic hierarchy implemented using such predicate classes. In sections~\ref{cat} and~\ref{univ} we give a taste of what category theory and universal algebra look like in our development, and in section~\ref{numbers} we use these facilities to build abstract interfaces for numbers. In order to illustrate a very different use of type classes, we discuss the implementation of a quoting function for algebraic terms in terms of type classes, in section~\ref{quoting}. In section~\ref{sequences} we hint at an interface for sequences, but describe how a limitation in the Coq system makes its use problematic. We end with conclusions in section~\ref{conclusions}.

@@

The Coq proof assistant is based on the calculus of inductive
constructions~\cite{CoquandHuet,CoquandPaulin}, a dependent type theory with (co)inductive types; see~\cite{Coq,BC04}. In true Curry-Howard fashion, it is both an excessively pure, if somewhat pedantic, functional programming language with an extremely expressive type system, and a language for mathematical statements and proofs. We highlight some aspect of Coq that are of particular relevance to our development.

\paragraph{Types and propositions.}

Propositions in Coq are types~\cite{ITT,CMCP}, which themselves have types called \emph{sorts}. Coq features a distinguished sort called \lstinline|Prop| that one may choose to use as the sort for types representing propositions. The distinguishing feature of the \lstinline|Prop| sort is that terms of non-\lstinline|Prop| type may not depend on the values of inhabitants of \lstinline|Prop| types (that is, proof terms).
% As originally conceived~\cite{ITT,CMCP}, Propositions are Types and we can extract information from
% them. Often we do \emph{not} want functions to depend on proofs. In Coq, this is the paradigmatic way of working. We are forbidden to extract information from proofs of \lstinline|Prop|s. Especially with the aid of the Program machinery, this allows us to flexibly work with simply typed program and their Hoare style correctness proofs.
This regime of discrimination establishes a weak form of proof irrelevance, in that changing a proof can never affect the result of value computations. On a very practical level, this lets Coq safely erase all \lstinline|Prop| components when extracting certified programs to OCaml or Haskell.
A stronger notion of proof irrelevance in which any two inhabitants of a \lstinline|Prop| type are made \emph{convertible} is planned for a future Coq version.
% todo: process?:
%One may be tempted to propose Propositions as [ ]-types~\cite{awodey2004propositions}. Here the
%construction $[P]:=\{\top\mid p:P\}$ squashes all the information about how we proved $P$. In Coq,
%the [ ]-types are isomorphic to the type of sort \lstinline|Prop|. However,
%there are places where we needed proof \emph{relevance}; e.g.\ in the proof of the first isomorphism
%theorem above.In practice, manual dead-code analysis seems useful. Making Harrop formulas proof
%irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. However, a further refinement
%of Propositions as types seems necessary.
% Making Harrop formulas proof irrelevant seems to be a good first approximation~\cite{lcf:spi:03}. An orthogonal and further reaching proposal is to replace the Coq type theory with the Calculus of implicit constructions with implicit Σ-types~\cite{miquel2001implicit,barras2008implicit,Bernardo}.

@@

The `native' notion of equality in Coq is that of terms being convertible, which in this context constitutes Leibniz equality, naturally propositionalized by an inductive type family \lstinline|eq| with a single constructor
\begin{lstlisting}
 eq_refl : Π (T: Type) (x: T), x ≡ x,
\end{lstlisting}
where `\lstinline|a ≡ b|' is notation for \lstinline|eq T a b|. Here we diverge from Coq tradition and reserve \lstinline|=| for setoid equality, as this is the equality we will be working with most of the time. Importantly, since convertibility is a congruence, a proof of \lstinline|a ≡ b| lets us substitute \lstinline|b| for \lstinline|a| anywhere inside a term without further conditions. We mention this explicitly only because rewriting \emph{does} give rise to conditions when we depart from Leibniz equality and introduce equivalence relations representing abstractions over common representations of the same conceptual object. For instance, consider formal fractions of integers as a representation of rationals. Rewriting a subterm using such an equality is permitted only if the subterm is an argument of a function that has been proven to \emph{respect} the equality. Such a function is called \emph{proper}, and propriety must be proved for each function in whose arguments we wish to enable rewriting.

Because the Coq type theory lacks quotient types (as it would make type checking undecidable), one usually bases abstract structures on a \emph{setoid} (`Bishop set'): a type equipped with an equivalence relation~\cite{Bishop67,Hofmann,Capretta}. As we will see in section~\ref{univ}, this way of working pays off when working with notions such as quotient algebras.

@@

The algebraic hierarchy of the ssreflect~\cite{Packed} libraries uses an interesting (if fairly restrictive) alternative approach. It requires unicity of representation for all objects so that Leibniz equality is always applicable.

\paragraph{Type classes.}

Type classes have been a great success story in the Haskell functional programming language, as a means of organizing interfaces of abstract structures. Coq's type classes provide a superset of their functionality, but implemented in a different way.

In Haskell and Isabelle, type classes and their instances are second class. They are handled as specialized syntactic constructs whose semantics are given specifically by the type class apparatus. By contrast, the expressivity of dependent types and inductive families as supported in Coq, combined with the use of pre-existing technology in the system (namely proof search and implicit arguments) enable a \emph{first class} type class implementation~\cite{DBLP:conf/tphol/SozeauO08}: classes are ordinary record types (`dictionaries'), instances are ordinary constants of these record types (registered as `hints' with the proof search machinery), class constraints are ordinary implicit parameters, and instance resolution is achieved by augmenting the unification algorithm to invoke ordinary proof search for implicit arguments of class type.
Thus, type classes in Coq are realized by relatively minor syntactic aids that bring together existing facilities of the theory and the system into a coherent idiom, rather than by introduction of a new category of qualitatively different definitions with their own dedicated semantics.

@@

More elaborate structures, too, can be expressed as predicates (expressing laws) over a number of carriers, relations, and operations. While optimally flexible in principle, in practice \emph{naive} adoption of this approach (that is, without using type classes) leads to substantial inconveniences in actual use: when \emph{stating} theorems about abstract instances of such structures, one must enumerate all components along with the structure (predicate) of interest. And when \emph{applying} such theorems, one must either enumerate any non-inferrable components, or let the system spawn awkward metavariables to be resolved at a later time. Importantly, this also hinders proof search for proofs of the structure predicates, making any nontrivial use of theorems a laborious experience. Finally, the lack of \emph{canonical names} for particular components of abstract structures makes it impossible to associate idiomatic notations with them.

@@

Manifest fields~\cite{DBLP:conf/types/Luo08} have been proposed, and implemented in the Matita system, to address exactly this problem. We hope to convince the reader that type system extensions of this nature, designed to mitigate particular symptoms of the bundled approach, are less elegant than a solution (described in the next section) that avoids the problem altogether by using predicate-like type classes in place of bundled records.

If we revert back to the predicate formulation of relations, we \emph{could} still define \lstinline|EquivalenceRelation| in a bundled fashion without the need for equalities:
\begin{lstlisting}
  Record EquivalenceRelation: Type :=
    { er_carrier: Type
    ; er_rel: relation er_carrier
    ; er_refl: ReflexiveRelation er_carrier er_rel
    ; er_sym: SymmetricRelation er_carrier er_rel
    ; er_trans: TransitiveRelation er_trans er_rel }.
\end{lstlisting}
However, as before we conclude that \mbox{\lstinline|EquivalenceRelation|,} too, should be a predicate. Indeed, it would be rather strange for the interface of equivalence relations to differ qualitatively from the interface of reflexive relations.

Another attempt to recover some grouping might be to bundle the carrier with the relation into a (lawless) record, but this too hinders sharing. As soon as we try to define an algebraic structure with two reflexive relations on the same carrier, we need awkward hacks to establish equality between the carrier projections of two different (carrier, relation) bundles.

Even bundling just the operations of an algebraic structure together in a record (with the carrier as a parameter) leads to the same problem when, for example, one attempts to define a hypothetical algebraic structure with two binary relations and a constant such that both binary relations form a monoid with the constant.

A second problem with bundling is that as the bundled records are stacked to represent higher and higher structures, the projection paths for their components grow longer and longer, resulting in ever more unwieldy terms (coercions and notations can make this less painful). Further, if one tries to implement some semblance of sharing in a bundled representation, these projection paths additionally become non-canonical, and still more extensions have been proposed to address this symptom, e.g.\ coercion pullbacks~\cite{Hints}.

@@

Next, we turn to problems concerning theorem \emph{declaration}. As a reference point, our ideal will be the common mathematical vernacular, where one simply says:
\begin{quote}
Theorem L: For $x, y, z$ in a semigroup $G$, $x * y * z = z * y * x$.
\end{quote}
(This silly statement allows us to clearly present the syntax.)

@@

\begin{lstlisting}
  Theorem L G e op {P: SemiGroup G e op}: Π x y z,
    equiv e (sg_op op (sg_op op x y) z) (sg_op op (sg_op op z y) x).
\end{lstlisting}
where \lstinline|e| and the \lstinline|op|s are filled in by instance resolution.

@@

Since \lstinline|CommutativeMonoid| indirectly includes a \lstinline|SemiGroup| field, which in turn includes a \lstinline|Equivalence| field, having a \lstinline|SemiRing| proof means having two distinct proofs that the equality relation is an equivalence. This kind of redundant knowledge ()which arises naturally) is never a problem in our setup, because the use of operational type classes ensures that terms composed of algebraic operations and relations never refer to structure proofs. We find that this is a tremendous relief compared to approaches that \emph{do} intermix the two and where one must be careful to ensure that such terms refer to the `right' proofs of properties. There, even \emph{strong} proof irrelevance (which would make terms convertible that differ only in what proofs they refer to) would not make these difficulties go away entirely, because high-level tactics that rely on quotation of terms require syntactic identity (rather than `mere' convertibility) to recognize identical subterms.

Because predicate classes only provide contextual information and are insulated from the actual algebraic expressions, their instances can always be kept entirely opaque---only their existence matters. Together, these properties largely defuse an argument occasionally voiced against type classes concerning perceived unpredictability of instance resolution. While it is certainly true that in contexts with redundant information it can become hard to predict which instance of a predicate class will be found by proof search, it simply \emph{does not matter} which one is found. Moreover, for operational type classes the issue rarely arises because their instances are not nearly as abundant.

@@

We use names for properties like distributivity and absorption, because these are type classes as well (which is why we declare them with \lstinline|:>|). It has been our experience that almost any generic predicate worth naming is worth representing as a predicate type class, so that its proofs will be resolved as instances behind the scenes whenever possible. Doing this consistently minimizes administrative noise in the code, bringing us closer to ordinary mathematical vernacular. Indeed, we believe that type classes are an elegant and apt formalization of the seemingly casual manner in which ordinary mathematical presentation assumes implicit administration and use of a `database' of properties previously proved. Much more so than the existing canonical structures facility, because canonical structures can only be used with bundled representations, which we argued against in section~\ref{bundling}.

The operational type classes used in \lstinline|SemiRing| for zero, one, multiplication and addition, are the same ones used by \lstinline|Ring| and \lstinline|Field| (not shown). Thus, the realization that a particular semiring is in fact a ring or field has no bearing on how one refers to the operations in question, which is as it should be. The realization that a particular semigroup is part of a semiring \emph{does} call for a new (canonical) name, simply because of the need for disambiguation. The introduction of these additional names for the same operation is quite harmless in practice, because canonical names established by operational type class fields are identity functions, so that in most contexts the distinction reduces away instantly.

@@

Some clarification is in order to explain the role of the \lstinline|Context| declaration of the two monoids. While \lstinline|Monoid_Morphism| seemingly depends on monoid-ness proofs (which would be a gross violation of our idiom), in fact it is only parameterized on the monoid \emph{components} declared through implicit generalization of the \lstinline|Monoid| declarations, because it only refers to those. Here, we use declarations of predicate class parameters merely as convenient shorthands to declare their components.

@@

Unfortunately, there is actually one annoying wrinkle here, which will also explain why we do not register these two fields as instance resolution hints (by declaring them with \lstinline|:>|). What we really want these fields to express is ``\emph{if} in a certain context we know something to be a \lstinline|Monoid_Morphism|, \emph{then} realize that the source and target are \lstinline|Monoid|s''. However, the current instance resolution implementation has little support for this style of \emph{forward} reasoning, and is really primarily oriented on \emph{backward} reasoning: had we registered \lstinline|monmor_to| and \lstinline|monmor_from| as instance resolution hints, we would in fact be saying ``\emph{if} trying to establish that something is a \lstinline|Monoid|, \emph{then} try finding a \lstinline|Monoid_Morphism| to or from it'', which quickly degenerates into a wild goose chase. We will return to this point in section~\ref{conclusions}.

@@

In Coq, for a term \lstinline|F| to function as though it had two different types simultaneously (namely the object map and the arrow map), there must either (1) be coercions from the type of F to either function, or (2) F must be (coercible to) a single function that is able to consume both object and arrow arguments. In addition to not being supported by Coq, option (1) would violate our policy of leaving components unbundled. For (2), if it could be made to work at all, F would need a pretty egregious type considering that arrow types are indexed by objects, and that the type of the arrow map
\begin{lstlisting}
 Π x y, (x ⟶ y) → (F x ⟶ F y)
\end{lstlisting}
must refer to the object map.

@@

To specify the natural numbers and the integers as initial objects in the categories of semirings and rings, respectively, definitions of these categories are needed. While one could define both of them manually, greater economy can be achieved by recognizing that both semirings and rings can be defined by equational theories, for which \emph{varieties} can be defined generically. Varieties are categories consisting of models for a fixed theory with homomorphisms between them.

To this end, we have formalized some of the theory of multisorted universal algebra and equational theories. We chose not to revive existing formalizations~\cite{DBLP:conf/tphol/Capretta99,dominguez2008formalizing} of universal algebra, because an important aim for us has been to find out what level of elegance, convenience, and integration can be achieved by leveraging the state of the art in Coq facilities (of which type classes are the most important example).

@@

Taking these conversion functions for granted, we will also need a ``down-to-earth'' representation of the initiality arrows if we are to give a \lstinline|SemiRing|-based definition of the interface for natural numbers. Once again, we introduce an operational type class to represent this particular component:
\begin{lstlisting}
  Class NaturalsToSemiRing (A: Type) :=
    naturals_to_semiring: Π B `{RingMult B} `{RingPlus B} `{RingOne B}
      `{RingZero B}, A → B.
\end{lstlisting}
To use this with \lstinline|Initial|, we define an additional conversion instance that takes a \lstinline|NaturalsToSemiRing| along with a proof showing that it yields \mbox{\lstinline|SemiRing_Morphism|s}, and builds an \lstinline|InitialArrow| instance out of it. This conversion instance in turn invokes another conversion function that translates concrete \lstinline|SemiRing_Morphism| proofs into univeral algebra \lstinline|Homomorphism|s instantiated with the semiring signature, which make up the arrows in the category.

@@

In our development we prove that the standard library's default rationals do indeed implement \lstinline|Rationals|, as do implementations of the \lstinline|QType| module interface. While the latter is rather ad-hoc from a theoretical perspective, it is nevertheless of great practical interest because it is used for the very efficient \lstinline|BigQ| rationals based on machine integers~\cite{machineintegers}. Hence, theory and programs developed on our \lstinline|Rationals| interface applies and can make immediate use of these efficient rationals. We plan to rebase the computable real number implementation~\cite{Oconnor:real} on this interface, precisely so that it may be instantiated with efficient implementations like these.
